Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-06-02 doc/manuscript_hkk_drought_sensitivity.Rmd
SJD basic text edits

Globally, droughts are becoming more frequent and intense, affecting forest demography, diversity, and  carbon (C) cycling [@ipcc_climate_2023]. 
However, our ability to predict drought impacts on tree woody growth and the resultant ecosystem C sequestration remains limited, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and through any drought. 
In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
<!-- Sean::: This paragraph is a bit redundant and all over the place. 1) in the tropics, seasonal dry periods (often defined as months with <= 100 mm rainfall) can have complex influences on tree phenology, from leaf turnover to growth allocation. Droughts, which can exacerbate or extend a dry season, or reduce cloud cover ina rainy season, would be expected interact with seasonal phenology also in complex ways. [Then go into some details with citations on ?] -->
Drought, or hot/dry conditions anomalously departing from climatic means during any time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019]. 
Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may in fact have very different effects on tree growth, especially in seasonal forests that have evolved phenological strategies to handle regular periods of climatic stress [@albert_ecology_letters_cryptic_phenology] [@gao_dynamic_2018].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
<!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology]. It seems like these arguments are a bit redundant. Or meandering. ---->
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those occurring at different times relative to regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

@@

We assigned deciduousness as a species level trait using data published in @williams_deciduousness_2008 for the top 30 species. 
We defined species deciduousness as the mean over trees of the proportion of crown leaf loss at maximum, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous.
Data was unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1, in the middle of the evergreen range. 
We also considered using the duration of time spent at maximum crown loss as an alternative measure of deciduousness.
This was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for more species.  
<!-- At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
We calculated niche breadth as the standard deviation of TWI across all observed individuals of each species in the whole plot census in 2008.  -->

@@

The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
The 2010 drought peaked between February and May (dry season and beginning of wet season), with 1-month SPEI values below -1 and meteorological anomalies on the 30-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions.
1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought.

In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum monthly mean temperatures (max anomaly +1.5 SD), and monthly maximum VPD (max anomaly +2 SD).
<!-- make sure my adjustments above in details are correct ... I think we should present a precise description  -->
1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census time period. 
Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season.
1- to 6-month SPEI values reached minima <2, indicative of extreme drought, while the 12-month values remained just above -2, indicative of severe drought.

@@

## Growth timeseries and drought responses {-}

The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a). 
Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
Median annual increment varied among species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed. 
<!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->

Drought years were associated with lower growth across the plot. 
In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years. 
The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years. 
While median species responses to the 2010 drought were variable, all species had lower median increment than a typical year in 2015. 
<!-- Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4).  -->
<!-- Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`.  -->
<!-- The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%. -->
<!-- *(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))* -->

On average, trees had negative drought sensitivities for both droughts, but with considerable variation (Fig `r fig_drought_responses`b). 
Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`. 
In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`*. 
In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`*.

@@

## Species effects {-}

On average, species has lower mean and median sensitivies in 2015 than in 2010 (Figure S5), in alignment with median sensitivities across all trees. 
Species median sensitivities in 2010were not correlated with median sensitivities in 2015 (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5).

Deciduousness species responded more negatively than evergreen species in 2015, but not in 2010 (Fig `r fig_species`a). 
In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`). 
In 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
Moreover, models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).

@@

TWI showed no significant effect on sensitivity in 2010, but did show a positive effect in 2015. 
Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`). 
The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
In 2015, evergreen trees responded more negatively in drier than in wetter areas, whereas deciduous trees - which were more negatively affected overall - responded more negatively in wetter than in drier areas. 
Combined, predicted sensitivities of deciduous trees were more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b).

The differing effects of TWI on sensitivity in the two droughts, along with the influence of deciduousness, were confirmed by models that controlled for crown exposure and DBH as well as species variation in intercept and slopes (Fig `r fig_microenvs`a). 
These models showed small negative sensitivities in wetter sites in 2010 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but a strong positive effect of TWI in 2015 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`). 
The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig `r fig_microenvs`b). 
Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not  increase with TWI (Fig `r fig_microenvs`c). 
On the other hand, neither the mean sensitivities across species nor their response to TWI were associated with deciduousness in 2010.

@@

DBH was a significant predictor of crown exposure, with larger DBH associated with higher exposure categories in 2010 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a). 
Controlling for this direct effect of exposure, the remaining effect of DBH *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).

@@

<!-- contributions / disentangling exposure from DBH-->
One variable examined here -- crown exposure -- had a consistent effect across the two droughts (Table 1, Figs. `r fig_microenvs`, `r fig_dag`). 
Trees with more exposed crowns exhibited greater growth declines in both droughts, while sub-canopy trees grew more than average during the 2010 drought.
This finding is consistent with the concept that larger trees tend to suffer more during drought [@bennett_larger_2015], but adds the important insight that it is crown position -- as opposed to size *per se* -- that is primarily responsible for this pattern.
The strong and consistent negative effect of crown position makes sense in light of pronounced gradients in drought stress across forest vertical profiles [@vinod_thermal_2023].
We show that residual effects of size, presumably including root water access and biophysical challenges linked to height itself, may act in different directions depending on the drought characteristics (Figs. `r fig_microenvs`, `r fig_dag`).
Together, these findings imply that while canopy and emergent trees face the greatest stress and exhibit the most negative growth responses in almost any drought, a suite of species traits and microenvironmental characteristics modify growth responses in interaction with drought characteristics.

>>>>

Intense droughts are becoming more frequent, affecting the demography, diversity, and  carbon (C) cycling of forests globally [@ipcc_climate_2023]. 
However, our ability to predict drought impacts on tree woody growth and ecosystem C sequestration remains limited, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and through any drought. 
In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
The complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
<!-- Sean::: This paragraph is a bit redundant and all over the place. 1) in the tropics, seasonal dry periods (often defined as months with <= 100 mm rainfall) can have complex influences on tree phenology, from leaf turnover to growth allocation. Droughts, which can exacerbate or extend a dry season, or reduce cloud cover ina rainy season, would be expected interact with seasonal phenology also in complex ways. [Then go into some details with citations on ?] -->
Drought, or hot/dry conditions anomalously departing from climatic means during any time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019]. 
Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may have very different effects on tree growth, especially in seasonal forests that have evolved phenological strategies to handle regular periods of climatic stress [@albert_ecology_letters_cryptic_phenology] [@gao_dynamic_2018].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
<!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology]. It seems like these arguments are a bit redundant. Or meandering. ---->
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those occurring at different times relative to regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

@@

We assigned deciduousness as a species level trait using data published in @williams_deciduousness_2008 for the top 30 species. 
We defined species deciduousness as the mean over trees of the proportion of crown leaf loss at maximum, with a score ranging from 0 to 4 where 0 is most evergreen and 4 is most deciduous.
Data were unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1. 
We also considered using the duration of time spent at maximum crown loss as an alternative measure of deciduousness.
This was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for more species.  
<!-- At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
We calculated niche breadth as the standard deviation of TWI across all observed individuals of each species in the whole plot census in 2008.  -->

@@

The two droughts analysed had distinct temporal characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
The 2010 drought peaked between February and May (dry season and beginning of wet season), with 1-month SPEI values below -1 and meteorological anomalies on the 30-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions.
1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought.

In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum monthly mean temperatures (max anomaly +1.5 SD), and monthly maximum VPD (max anomaly +2 SD).
<!-- make sure my adjustments above in details are correct ... I think we should present a precise description  -->
1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period. 
Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season.
1- to 6-month SPEI values reached minima <2, indicative of extreme drought, while the 12-month values remained just above -2, indicative of severe drought.

@@

## Growth responses to drought {-}

<!-- The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a).  -->
Median annual growth rate across all years of measurement for the whole plot was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
In the 2010 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years. 
The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years. 
On average, trees had negative drought sensitivities for both droughts, but with considerable variation (Fig `r fig_drought_responses`b). 
Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`. 
<!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->

<!-- Drought years were associated with lower growth across the plot.  -->
<!-- Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4).  -->
<!-- Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`.  -->
<!-- The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%. -->
<!-- *(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))* -->

@@

## Species and functional group drought responses {-}

Median annual increment varied among species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed.  
While median species responses to the 2010 drought were variable, all species had lower median increment than a typical year in 2015. 
In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`*. 
In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`*.

Species median sensitivities to the 2010 drought were not correlated with median sensitivities in 2015 (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5) but on average, species had more negative sensitivities to the 2015 drought than the 2010 drought, in alignment with median sensitivities across all trees.

<!-- Deciduousness species responded more negatively than evergreen species in 2015, but not in 2010 (Fig `r fig_species`a).  -->
Predicted sensitivities from intercept-only models with a species random effect showed that  deciduous species were more drought sensitive than evergreen species in 2015 (Fig `r fig_species`a, r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`), but not in 2010 (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
Models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).

@@

TWI showed no significant effect on sensitivity in 2010, but did show a positive effect in 2015. 
Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`). 
The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
In 2015, evergreen trees responded more negatively in drier than in wetter areas, whereas deciduous trees - which were more negatively affected overall - responded more negatively in wetter than in drier areas. 
Predicted sensitivities of deciduous trees were more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b).

The differing effects of TWI on sensitivity in the two droughts, along with the influence of deciduousness, were confirmed by models that controlled for crown exposure and DBH as well as species variation in intercept and slopes (Fig `r fig_microenvs`a). 
These models showed small negative sensitivities in wetter sites in 2010 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but a strong positive effect of TWI in 2015 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`). 
The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig `r fig_microenvs`b). 
Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not  increase with TWI (Fig `r fig_microenvs`c). 
Mean sensitivities across species and their response to TWI were not associated with deciduousness in 2010.

@@

<!--DBH was a significant predictor of crown exposure, -->Trees with larger DBH had higher crown exposure in 2010 (median effect = `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect = `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a). 
Controlling for the direct effect of exposure, the remaining effect of DBH *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).

@@

<!-- contributions / disentangling exposure from DBH-->
<!-- One variable examined here -- crown exposure -- had a consistent effect across the two droughts (Table 1, Figs. `r fig_microenvs`, `r fig_dag`).  -->
Trees with more exposed crowns exhibited greater growth declines in both droughts, while sub-canopy trees grew more than average during the 2010 drought.
This finding is consistent with the concept that larger trees tend to suffer more during drought [@bennett_larger_2015], but adds the important insight that it is crown position -- as opposed to size *per se* -- that is primarily responsible for this pattern.
The strong and consistent negative effect of crown position makes sense in light of pronounced gradients in drought stress across forest vertical profiles [@vinod_thermal_2023].
We show that residual effects of size, presumably including root water access and biophysical challenges linked to height itself, may act in different directions depending on the drought characteristics (Figs. `r fig_microenvs`, `r fig_dag`).
Together, these findings imply that while canopy and emergent trees face the greatest stress and exhibit the most negative growth responses in almost any drought, a suite of species traits and microenvironmental characteristics modify growth responses in interaction with drought characteristics.
